---
title: "Modelling the effects of public health interventions on COVID-19 transmission using R - part 2"
description: |
  This is the secondof several posts which examine the effect of various public health interventions on the local epidemic spread of COVID-19 infection using stochastic compartmental models implemented by the `EpiModel` library for R. In this post we extend the ICM SIR model provided by `EpiModel` in various ways to improve its verisimilitude.
categories:
  - R
  - "COVID-19"
author:
  - name: Tim Churches 
    affiliation: South Western Sydney Clinical School, UNSW Medicine & Ingham Institute of Applied Medical Research, Liverpool, Sydney, Australia
    affiliation_url: https://swscs.med.unsw.edu.au
creative_commons: CC BY-SA
date: 03-11-2020
base_url: https://timchurches.github.io/blog/
slug: churches2020modellingcovid19rpart2
twitter:
  creator: "@timchurches"
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    self_contained: false
draft: true
repository_url: https://github.com/timchurches/blog/tree/master/_posts/2020-03-11-modelling-the-effects-of-public-health-interventions-on-covid-19-transmission-part-2
bibliography: churches2020modellingcovid19rpart2_references.bib
---


```{r setup, include=FALSE, eval=TRUE}
version_date <- lubridate::ymd("2020-03-11")

knitr::opts_chunk$set(echo = FALSE, cache=FALSE, eval=TRUE,
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE)
library(tidyverse)
library(magrittr)
library(lubridate)
library(stringr)
library(tibble)
library(broom)
library(ggplot2)
library(ggthemes)
library(gt)
library(knitr)
library(devtools)
library(DiagrammeR)
library(parallel)
library(foreach)
suppressMessages(library(EpiModel))

```

# Introduction

We'll extend the SIR model quite a bit, adding the following compartments (states):

| Compartment | Functional definition                            |
|=============|==================================================|
| S           | Susceptible individuals                          |
| E           | Exposed **and** infected, but not yet infectious |
| I           | Infected **and** infectious                      |
| Q           | Infectious, by quarantined or (self-)isolated    |
| H           | Requiring hospitalisation (would normally be hospitalised if capacity avaialible) |
| R           | Recovered, immune from further infection         |
| F           | Case fatality (death due to COVID-19, not other causes) |

The permitted transitions between compartments are as shown in this diagram:

```{r, echo=FALSE, eval=TRUE, message=FALSE}
grViz("
digraph SEIQHRF {

  # a 'graph' statement
  graph [overlap = false, fontsize = 10] #, rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  S[label='S=Susceptible'];
  E[label='E=Exposed and infected,\nnot infectious'];
  I[label='I=Infected and infectious'];
  Q[label='(Self-)isolated/quarantined'];
  H[label='Requires hospitalisation'];
  R[label='Recovered/immune'];
  F[label='Case fatality']

  # several 'edge' statements
  S->E
  I->S[style='dotted']
  E->I
  I->Q
  Q->S[style='dotted']
  I->R
  I->H
  H->F
  H->R
  Q->R
  Q->H
}
")
```

```{r, echo=FALSE, eval=FALSE, message=FALSE}
grViz("
digraph SEIQHRF {

  # a 'graph' statement
  graph [overlap = false, fontsize = 10] #, rankdir = LR]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica]
  S[label='Susceptible'];
  E[label='Infected,\nnot infectious'];
  I[label='Infectious'];
  Q[label='Isolated/quarantined'];
  H[label='Requires hospitalisation'];
  R[label='Recovered/immune'];
  F[label='Case fatality']

  # several 'edge' statements
  S->E
  I->S[style='dotted', label='act.rate.i\ninf.prob.i']
  E->I[label='prog.rate\nprog.dist.mu\nprog.dist.sigma']
  I->Q
  Q->S[style='dotted', label='act.rate.q\ninf.prob.q']
  I->R
  I->H
  H->F
  H->R
  Q->R
  Q->H
}
")
```

## SEIQHRF model with gamma progress and recovery probs

For now, we'll just monkey-patch the EpiModel code with our extensions, which can be pulled from a GitHub gist.  Later we'll provide a proper package.

```{r, echo=FALSE, eval=TRUE, message=TRUE}
source_files <- c("_icm.mod.init.seiqhrf.R",
                  "_icm.mod.status.seiqhrf.R",
                  "_icm.mod.vital.seiqhrf.R",
                  "_icm.control.seiqhrf.R",
                  "_icm.utils.seiqhrf.R",
                  "_icm.saveout.seiqhrf.R",
                  "_icm.icm.seiqhrf.R")
src_path <- "./_posts/2020-03-11-modelling-the-effects-of-public-health-interventions-on-covid-19-transmission-part-2/"
gist_url <- "https://gist.github.com/timchurches/92073d0ea75cfbd387f91f7c6e624bd7"

local_source <- TRUE
for (source_file in source_files) {
  if (local_source) {
    source(paste(src_path, source_file, sep=""))
  } else {
    source_gist(gist_url, filename=source_file)
  }
}
```

We'll define a function called `simulate()` that wraps the various simulation constructors and provides baseline defaults for the (very) many parameters. You can change the defaults in this function definition for you own model comparisons.

Public health interventions can then be investigated by changing just those parameters which relate to or represent the intervention. 

The available parameters are as follows:

```{r, echo=FALSE, eval=FALSE, message=FALSE}
param_docs <- tribble(~name, ~default, ~explanation,
 "type", "SEIQHRF", "Type of model: SI, SIR, SIS, SEIR, SEIQHR and SEIQHRF available, only SEIQHRF fully tested.",

  "nsteps", 366, "Number of days for simulation. Note that day 1 is for initialisation, day 2 is the first day of the simulation, hence default of 366 for 1 year.",

  "nsims", 10, "Number of simulations to run and then average.",

  "prog.rand", FALSE, "Method for progression from E to I. If TRUE, random binomial draws at `prog.rate`, if FALSE, random draws from a truncated gamma distribution, with mean `prog.dist.mu` and standard deviation `prog.dist.sigma`",

  "rec.rand", FALSE, "Method for recovery transition from I, Q or H to R. If TRUE, random binomial draws at `prog.rate`, if FALSE, random draws from a truncated gamma distribution, with mean `prog.dist.mu` and standard deviation `prog.dist.sigma`",

  "fat.rand", FALSE, "Method for case fatality transition from H to F. If TRUE, random binomial draws at `prog.rate`, if FALSE, random draws from a uniform distribution, with a rate given by `fat.rate.base`. However, if the current number of patients in the H (needs hospitalsiation) compartment is above a hospital capacity level specified by `hosp.cap`, then the fatality rate is a weighed sum of the base fatality rate up to the hospital capacity, then at a higher rate, specified by `fat.rate.overcap`.",

  "quar.rand", FALSE, "Method for quarantine transition from I to Q. If TRUE, random binomial draws at `quar.rate`, if FALSE, random draws from a uniform distribution, with a rate also given by `quar.rate.",

  "hosp.rand", FALSE, "Method for transition from I or Q to H -- that is, from infectious or from quarantined/isolated to requiring hospitalisation. If TRUE, random binomial draws at `hosp.rate`, if FALSE, random draws from a uniform distribution, with a rate also given by `hosp.rate.",
  
 )
  
kable(param_docs)
```

```
 quar.rand = TRUE,
 hosp.rand = FALSE,
 disch.rand = TRUE,
 infection.FUN = infection.seiqhrf.icm,
 recovery.FUN = progress.seiqhrf.icm,
 departures.FUN = departures.seiqhrf.icm,
 # arrivals.FUN = arrivals.seiqhrf.icm,
 arrivals.FUN = arrivals.icm,
 get_prev.FUN = get_prev.seiqhrf.icm,
 # init.icm params
 s.num = 997,
 e.num=0,
 i.num = 3,
 q.num=0,
 h.num=0,
 r.num = 0,
 f.num = 0,
 # param.icm params
 inf.prob.i = 0.05, 
 act.rate.i = 10,
 inf.prob.q = 0.05, 
 act.rate.q = 8,                    
 quar.rate = 0, # 1/3
 hosp.rate = 1/50,
 disch.rate = 1/12,
 prog.rate = 1/10,
 prog.dist.mu = 8.0,
 prog.dist.sigma = 3,
 rec.rate = 1/20,
 rec.dist.mu = 22.0,
 rec.dist.sigma = 6.0,
 fat.rate.base = 1/100,
 hosp.cap = 20,
 fat.rate.overcap = 1/50,
 fat.tcoeff = 0.5,
 vital = TRUE,
 a.rate = (10.5/365)/1000, 
 a.prop.e = 0.01,
 a.prop.i = 0.001,
 a.prop.q = 0.01,
 ds.rate = (7/365)/1000, 
 de.rate = (7/365)/1000, 
 di.rate = (7/365)/1000,
 dq.rate = (7/365)/1000,
 dh.rate = (20/365)/1000,
 dr.rate = (7/365)/1000,
 out="mean"
```

```{r, echo=TRUE, eval=TRUE}
# function to set-up and run the baseline simulations
simulate <- function(# control.icm params
                     type = "SEIQHRF", 
                     nsteps = 366, 
                     nsims = 8,
                     ncores = 4,
                     prog.rand = FALSE,
                     rec.rand = FALSE,
                     fat.rand = TRUE,
                     quar.rand = FALSE,
                     hosp.rand = FALSE,
                     disch.rand = TRUE,
                     infection.FUN = infection.seiqhrf.icm,
                     recovery.FUN = progress.seiqhrf.icm,
                     departures.FUN = departures.seiqhrf.icm,
                     # arrivals.FUN = arrivals.seiqhrf.icm,
                     arrivals.FUN = arrivals.icm,
                     get_prev.FUN = get_prev.seiqhrf.icm,
                     # init.icm params
                     s.num = 997,
                     e.num=0,
                     i.num = 3,
                     q.num=0,
                     h.num=0,
                     r.num = 0,
                     f.num = 0,
                     # param.icm params
                     inf.prob.e = 0.02, 
                     act.rate.e = 10,
                     inf.prob.i = 0.05, 
                     act.rate.i = 10,
                     inf.prob.q = 0.02, 
                     act.rate.q = 2.5,                    
                     quar.rate = 1/10, #1/3,
                     hosp.rate = 1/100,
                     disch.rate = 1/15,
                     prog.rate = 1/10,
                     prog.dist.scale = 5,
                     prog.dist.shape = 1.5,
                     rec.rate = 1/20,
                     rec.dist.scale = 35,
                     rec.dist.shape = 1.5,
                     fat.rate.base = 1/50,
                     hosp.cap = 40,
                     fat.rate.overcap = 1/25,
                     fat.tcoeff = 0.5,
                     vital = TRUE,
                     a.rate = (10.5/365)/1000, 
                     a.prop.e = 0.01,
                     a.prop.i = 0.001,
                     a.prop.q = 0.01,
                     ds.rate = (7/365)/1000, 
                     de.rate = (7/365)/1000, 
                     di.rate = (7/365)/1000,
                     dq.rate = (7/365)/1000,
                     dh.rate = (20/365)/1000,
                     dr.rate = (7/365)/1000,
                     out="mean"
                    ) {

  control <- control.icm(type = type, 
                         nsteps = nsteps, 
                         nsims = nsims,
                         ncores = ncores,
                         prog.rand = prog.rand,
                         rec.rand = rec.rand,
                         infection.FUN = infection.FUN,
                         recovery.FUN = recovery.FUN,
                         arrivals.FUN = arrivals.FUN,
                         departures.FUN = departures.FUN,
                         get_prev.FUN = get_prev.FUN)

  init <- init.icm(s.num = s.num,
                   e.num = e.num,
                   i.num = i.num,
                   q.num = q.num,
                   h.num = h.num,
                   r.num = r.num,
                   f.num = f.num)

  param <-  param.icm(inf.prob.e = inf.prob.e, 
                      act.rate.e = act.rate.e,
                      inf.prob.i = inf.prob.i, 
                      act.rate.i = act.rate.i,
                      inf.prob.q = inf.prob.q, 
                      act.rate.q = act.rate.q,                    
                      quar.rate = quar.rate,
                      hosp.rate = hosp.rate,
                      disch.rate = disch.rate,
                      prog.rate = prog.rate,
                      prog.dist.scale = prog.dist.scale,
                      prog.dist.shape = prog.dist.shape,
                      rec.rate = rec.rate,
                      rec.dist.scale = rec.dist.scale,
                      rec.dist.shape = rec.dist.shape,
                      fat.rate.base = fat.rate.base,
                      hosp.cap = hosp.cap,
                      fat.rate.overcap = fat.rate.overcap,
                      fat.tcoeff = fat.tcoeff,
                      vital = vital,
                      a.rate = a.rate, 
                      a.prop.e = a.prop.e,
                      a.prop.i = a.prop.i,
                      a.prop.q = a.prop.q,
                      ds.rate = ds.rate, 
                      de.rate = de.rate, 
                      di.rate = di.rate,
                      dq.rate = dq.rate,
                      dh.rate = dh.rate,
                      dr.rate = dr.rate)

  sim <- icm.seiqhrf(param, init, control)
  sim_df <- as.data.frame(sim, out=out)

  return(list(sim=sim, df=sim_df))
}
```

```{r, echo=TRUE, eval=TRUE}
#library(profvis)
#profvis({
  seiqhrf_sim <- simulate(ncores=4)
#})
```

Let's examine the distributions of timings for various transitions.

```{r, echo=TRUE, eval=TRUE}
# define a function to extract timings and assemble a data frame
get_times <- function(simulate_results) {

  sim <- simulate_results$sim
  
  for (s in 1:sim$control$nsims) {
    if (s == 1) {
      times <- sim$times[[paste("sim",s,sep="")]]
      times <- times %>% mutate(s=s)
    } else {
      times <- times %>%
        bind_rows(sim$times[[paste("sim",s,sep="")]] %>%
                    mutate(s=s))
    }
  }
  
  times <- times %>%
            mutate(infTime=ifelse(infTime <0, -5, infTime),
                   expTime=ifelse(expTime <0, -5, expTime)) %>%
            mutate(incubation_period = infTime - expTime,
                   illness_duration = recovTime - expTime,
                   illness_duration_hosp = dischTime - expTime,
                   hosp_los = dischTime - hospTime,
                   quarantine_delay = quarTime - infTime,
                   survival_time = fatTime - infTime) %>%
            select(s, 
                   incubation_period,
                   quarantine_delay,
                   illness_duration,
                   illness_duration_hosp,
                   hosp_los,
                   survival_time) %>%
            pivot_longer(-s, names_to="period_type",
                             values_to="duration") %>%
            mutate(period_type = factor(period_type, levels=c("incubation_period",
                                            "quarantine_delay",
                                            "illness_duration",
                                            "illness_duration_hosp",
                                            "hosp_los",
                                            "survival_time"),
                                   labels=c("Incubation period",
                                            "Delay entering quarantine",
                                            "Illness duration",
                                            "Illness duration (hosp)",
                                            "Hospital care required duration",
                                            "Survival time of case fatalities"),
                                   ordered = TRUE))
  return(times)
}
```

```{r, echo=TRUE, eval=TRUE}

times <- get_times(seiqhrf_sim)

```

```{r, echo=TRUE, eval=TRUE, fig.height=12, fig.width=8}
times %>%
  filter(duration <= 30) %>%
  ggplot(aes(x=duration)) +
    geom_bar() +
    facet_grid(period_type~., scales="free_y") +
    labs(title="Duration frequency distributions")
```


```{r, echo=TRUE, eval=TRUE, fig.height=12, fig.width=10}
seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "SEIQHRF") %>%
  filter(time <= 100) %>%
  pivot_longer(-c(time, model),
               names_to="compartment",
               values_to="count") %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.7) +
    scale_colour_brewer(type="div") +
    theme_dark()
```

```{r, echo=TRUE, eval=TRUE, fig.height=12, fig.width=10}
seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num,
                     h.num, r.num, f.num) %>%
              mutate(model = "SEIQHRF") %>%
  filter(time <= 100) %>%
  pivot_longer(-c(time, model),
               names_to="compartment",
               values_to="count") %>%
  filter(compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.7) +
    scale_colour_brewer(type="div") +
    theme_dark()
```

## comparing SEIQHRF models - fixed vs time-variant act rates and inf probs

```{r, echo=TRUE, eval=TRUE}
t30_raise_hosp_cap_seiqhrf_sim <- simulate(hosp.cap = c(rep(8,30), rep(14, 335)))

t30_act_rate_seiqhrf_sim <- simulate(act.rate.i = c(rep(10,30), rep(5, 335)),
                                     act.rate.e = c(rep(10,30), rep(5, 335)))

t30_inf_prob_seiqhrf_sim <- simulate(inf.prob.i = c(rep(0.05,30), rep(0.025, 335)),
                                     inf.prob.e = c(rep(0.02,30), rep(0.01, 335)))

t30_both_seiqhrf_sim <- simulate(inf.prob.i = c(rep(0.05,30), rep(0.025, 335)),
                                    act.rate.i = c(rep(10,30), rep(5, 335)),
                                 inf.prob.e = c(rep(0.02,30), rep(0.01, 335)),
                                    act.rate.e = c(rep(10,30), rep(5, 335)))

t30_quar_inc_seiqhrf_sim <- simulate(quar.rate = c(rep(1/10,30), rep(1/3, 335)))

t15_act_rate_seiqhrf_sim <- simulate(act.rate.i = c(rep(10,15), rep(5, 350)),
                                     act.rate.e = c(rep(10,15), rep(5, 350)))

t15_inf_prob_seiqhrf_sim <- simulate(inf.prob.i = c(rep(0.05,15), rep(0.025, 350)),
                                     inf.prob.e = c(rep(0.02,15), rep(0.01, 350)))

t15_both_seiqhrf_sim <- simulate(inf.prob.i = c(rep(0.05,15), rep(0.025, 350)), 
                                    act.rate.i = c(rep(10,15), rep(5, 350)),
                                 inf.prob.e = c(rep(0.02,15), rep(0.01, 350)), 
                                    act.rate.e = c(rep(10,15), rep(5, 350)))
t15_both_i_only_seiqhrf_sim <- simulate(inf.prob.i = c(rep(0.05,15), rep(0.025, 350)), 
                                    act.rate.i = c(rep(10,15), rep(5, 350)))
                                 
```

```{r, echo=TRUE, eval=TRUE, fig.height=12, fig.width=6}
bind_sim_rows <- function(sim_df_a, sim_df_b, label=NULL) {
  sim_df_c <- sim_df_a %>%
                bind_rows(sim_df_b %>%
                    select(time, s.num, e.num, i.num, q.num, 
                            h.num, r.num, f.num) %>%
                    mutate(model = label))
  return(sim_df_c)
}
                

seiqhrf_sim$df %>%
  select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "Fixed params") %>%
  bind_rows(t30_inf_prob_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=30 inf prob")) %>%
  bind_rows(t30_act_rate_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=30 act rate")) %>%
  bind_rows(t30_both_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=30 both")) %>%
  bind_rows(t30_quar_inc_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=30 incr quar rate")) %>%
  bind_rows(t30_raise_hosp_cap_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=30 hosp cap")) %>%
  bind_rows(t15_act_rate_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=15 act rate")) %>%
  bind_rows(t15_inf_prob_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=15 inf prob")) %>%
  bind_rows(t15_both_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=15 both")) %>%
  bind_rows(t15_both_i_only_seiqhrf_sim$df %>%
              select(time, s.num, e.num, i.num, q.num, 
                     h.num, r.num, f.num) %>%
              mutate(model = "t=15 both (i)")) %>%
filter(time <= 150) %>%
  pivot_longer(-c(time, model),
               names_to="compartment",
               values_to="count") %>%
  filter(compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) -> plot_df

plot_df %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.4) +
    facet_grid(model ~ .) +
    scale_colour_viridis_d() +
    theme_dark()
```

```{r, echo=TRUE, eval=TRUE, fig.height=12, fig.width=6}
plot_df %>%
  filter(compartment %in% c("h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.4) +
    facet_grid(model ~ .) +
    scale_colour_viridis_d() +
    theme_dark()
```

Up to here

```{r, echo=TRUE, eval=FALSE, fig.height=12, fig.width=10}
seir_sim_df %>%
  select(time, s.num, e.num, i.num, r.num) %>%
  mutate(model = "SEIR") %>%
  bind_rows(sir_sim_df %>%
              select(time, s.num, i.num, r.num) %>%
              mutate(model = "SIR")) %>%
  bind_rows(gamma_seir_sim_df %>%
              select(time, s.num, e.num, i.num, r.num) %>%
              mutate(model = "gamma SEIR")) %>%
  bind_rows(seiqhr_sim_df %>%
              select(time, s.num, e.num, i.num, q.num,
                     h.num, r.num, dh.flow) %>%
              mutate(model = "SEIQHR")) %>%
  bind_rows(seiqhrf_sim_df %>%
              select(time, s.num, e.num, i.num, q.num,
                     h.num, r.num, f.num) %>%
              mutate(model = "SEIQHRF")) %>%
  filter(time <= 100) %>%
  pivot_longer(-c(time, model),
               names_to="compartment",
               values_to="count") %>%
  filter(compartment %in% c("e.num","i.num",
                            "q.num","h.num",
                            "f.num")) %>%
  ggplot(aes(x=time, y=count, colour=compartment)) +
    geom_line(size=2, alpha=0.4) +
    facet_grid(model ~ ., scales="free_y") +
    scale_colour_brewer(type="div") +
    theme_minimal()
```
