---
title: "COVID-19 trends and trajectories"
description: |
  A short description of the post.
categories:
  - R
  - "COVID-19"
author:
  - name: Tim Churches 
    affiliation: South Western Sydney Clinical School, UNSW Medicine & Ingham Institute of Applied Medical Research, Liverpool, Sydney, Australia
    affiliation_url: https://swscs.med.unsw.edu.au
  - name: Nick Tierney 
    affiliation: Monash University
    affiliation_url: 
creative_commons: CC BY-NC-SA
date: 03-20-2020
base_url: https://timchurches.github.io/blog/
slug: churches2020covid19trendsandtrajectories
twitter:
  creator: "@timchurches"
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    self_contained: true
draft: true
repository_url: https://github.com/timchurches/blog/tree/master/_posts/2020-03-20-covid-19-trends-and-trajectories
---

```{r setup, include=FALSE}
version <- "1.8"
version_date <- lubridate::ymd("2020-02-28")

knitr::opts_chunk$set(echo = FALSE, cache=FALSE,
                      tidy.opts=list(width.cutoff=60),
                      tidy=TRUE)

library(tidyverse)
library(magrittr)
library(lubridate)
library(tibble)
library(ggplot2)
library(ggthemes)
library(hrbrthemes)
library(rvest)
library(gt)
library(deSolve)
library(EpiEstim)
library(incidence)
library(distcrete)
library(epitrix)
library(projections)
library(ggiraph)
```

```{r}
  provinces_confirmed_jh <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_19-covid-Confirmed.csv") %>%
    rename(province="Province/State",
           country_region="Country/Region") %>%
    pivot_longer(-c(province, country_region, Lat, Long), 
                 names_to = "Date",
                 values_to="cumulative_cases") %>%
    mutate(Date=mdy(Date)) %>%
    arrange(country_region, province, Date) %>%
    group_by(country_region, province) %>%
    mutate(incident_cases = c(0,diff(cumulative_cases))) %>%
    ungroup()

  incidence_countries <- provinces_confirmed_jh %>%
    group_by(country_region, Date) %>%
    summarise(incident_cases = sum(incident_cases),
              cumulative_cases = sum(cumulative_cases))
 
    incidence_global <- provinces_confirmed_jh %>%
    group_by(Date) %>%
    summarise(incident_cases = sum(incident_cases),
              cumulative_cases = sum(cumulative_cases))
 
```

```{r}
# find dates on which 100 cumulative cases occurred
high_incidence_countries <- incidence_countries %>%
  arrange(country_region, Date) %>%
  mutate(lag_cum_cases = lag(cumulative_cases)) %>%
  mutate(hit_100 = ifelse((cumulative_cases >= 100) & 
                          (lag(cumulative_cases) < 100),
                          TRUE,
                          FALSE))

high_incidence_countries <- high_incidence_countries %>% 
    left_join(high_incidence_countries %>% 
      group_by(country_region) %>%
      summarise(over_100 = any(hit_100))) %>%
  ungroup() %>%
  filter(over_100 == TRUE)

high_incidence_countries <- high_incidence_countries %>%
  arrange(country_region, Date) %>%
  group_by(country_region) %>%
  mutate(rnum = row_number()) %>%
  ungroup()

highlight_countries <- c("Japan", "Korea, South",
                          "Singapore", "Italy",
                          "Iran", "China", "France",
                          "United Kingdom",
                          "Sweden", "Australia")

high_incidence_countries <- high_incidence_countries %>%
  left_join(high_incidence_countries %>% 
              filter(hit_100) %>%
              mutate(offset = rnum) %>%
              select(country_region, offset)) %>%
  mutate(normalised_date = rnum - offset) %>%
  select(-c(lag_cum_cases, over_100, offset)) %>%
  mutate(alpha = case_when(country_region == "Australia" ~ 1.0,
                           country_region %in% highlight_countries ~ 0.2,
                           TRUE ~ 0)) %>%
  arrange(country_region, normalised_date) %>%
  group_by(country_region) %>%
  mutate(clabel_value = ifelse(country_region %in% highlight_countries,
                         country_region,
                         NA),
         clabel_x = max(normalised_date),
         clabel_y = last(cumulative_cases))

```

## The "Grattan Institute" plot

```{r, layout="l-page"}
gg <- high_incidence_countries %>%
  filter(normalised_date >= 0,
         country_region != "Cruise Ship") %>%
  ggplot(aes(x=normalised_date, y=cumulative_cases, 
             colour=country_region, alpha=alpha, 
             tooltip=country_region,
             data_id=country_region)) +
  geom_line_interactive(size=1.2) + 
  geom_label(aes(x=clabel_x, y=clabel_y, label=clabel_value),
             hjust = 0, nudge_x = 0.2) +
  scale_y_log10(labels=scales::comma) +
  scale_alpha(range=c(0.2, 1)) +
  expand_limits(x=c(0, max(high_incidence_countries$normalised_date) + 3)) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(y="Cumulative cases (logarithmic scale)",
       x="Days since cumulative cases exceeded 100",
       title=paste("COVID-19 epidemic trajectories up to", format(max(high_incidence_countries$Date), "%d %B %Y")),
       subtitle="(click on a line to identify country)",
       caption=paste0("CC-BY-SA-NC Tim Churches (UNSW) & Nick Tierney (Monash)\n",
                "Data source: Johns Hopkins University"))

girafe(ggobj=gg,
       options = list(opts_tooltip(use_stroke = TRUE, opacity=0.7),
                      opts_hover(css = "stroke:black;"),
                      opts_toolbar(position = "topright", saveaspng = TRUE)
                      ))
```

## Instantaneous effective reproduction number

```{r, eval=FALSE, echo=FALSE}
hubei_incidence_function_data <- provinces_confirmed %>%
      filter(Date >= ymd("2020-01-11")) %>% 
      mutate(HubeiSansWuhan = ifelse(is.na(HubeiSansWuhan), 
                                      0, HubeiSansWuhan)) %>%
      mutate(incident_cases = ifelse(Date < ymd("2020-02-15"),
                                     Wuhan + HubeiSansWuhan, Hubei)) %>% mutate(Date = format(Date, 
    "%Y-%m-%d")) %>% 
    select(Date, incident_cases) 
    %>% uncount(incident_cases)

hubei_incidence_object <- incidence(hubei_incidence_function_data$Date)
